Mechanical loading prediction though accelerometry data during walking and running

### 2021-06-09 manuscript/manuscript.Rmd
Add more HF suggetions

@@

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and acceleration (ACC; expressed in gravitational ACC units; 1*g* = 9.807 m·s^-2^) from the x, y and z vectors were exported to plain text files.

@@

FP and accelerometer data were processed using MATLAB (Version 2019a, Mathworks, Natick, USA), as described elsewhere [@Veras2020]. Briefly, both GRF and acceleration signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, and their resultant vector calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, peak ACC (pACC) values were defined as minimum height of twice the positive ACC average, during any given speed, and separated by 0.4 seconds minimum. GRF and ACC signals were adjusted through the time set by the systems clock, with manual correction based on visual inspection for possible errors, and then synchronized by the maximum cross-correlation coefficient. Peak GRF (pGRF) was determined as the highest value between 0.4 seconds before and after each pACC. The rates of change were computed through a centered derivative from the beginning of the foot contact to the pGRF. The peak loading rate (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array.

@@

Finally, our pGRF prediction equation was compared with a previously published reference equation [@Neugebauer2014], that was developed using similar methods. This comparison was performed in three ways using: i) the whole sample; ii) a subsample of walking trials; iii) a subsample of running trials. To assess the prediction accuracy, MAE, MAPE and RMSE were calculated using pGRF values predicted from both equations. No pLR prediction equation suited to be compared with ours was found.

@@

Furthermore, we compared the performance of our equations with a reference equation previously published. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, a sub-analysis was done separating walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all better compared to the reference equation. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy, which cannot be observed in the reference equation.

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, it can be seen that prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\% but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no cross-validation, which is similar to our results when no cross-validation method is applied (data not shown). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that distinguishes whether the activity is walking or running, which is not easy to be identify from free-living accelerometer data, hindering the real-world application of these previous equations. Also, one of these studies included in their sample only children and adolescents [@Neugebauer2012], that present a higher variability in their gait pattern compared to the adults [@Gouelle2016], making this equation unsuitable to use in an adult population. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. Accordingly to what we found, these studies also showed that the prediction error for these models is considerably larger than for the pGRF prediction. One of these researches investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], while the other obtained values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in these previous investigations have both lower and higher prediction errors compared to our results, which were approximately of 35\%. Although pLR is an important variable, known for a long time to determine to a great extent the bone formation response to mechanical loading [@Bassey1998; @Turner1995], based on our findings and in previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data does not seem to yield trustworthy results. Nevertheless, past research has suggested that the bone response to pGRF and pLR is somewhat similar during walking, running and jumping activities due to the high correlation between these variables [@Heikkinen2007]. Therefore, the pGRF prediction by accelerometry data, which is shown to be accurate, already incorporates some of the LR effect on the bone response to loading.

Results from our study offer some advantages compared to previous investigations: i) it includes walking and running, which are among the most usual types of physical activity performed daily; ii) all models developed, especially for pGRF prediction, presented a similar or better accuracy compared to the results from previous publications; iii) the accuracy of our models is constant throughout walking and running, notably for the pGRF prediction models, and therefore they can be used adequately for both activities; iv) in order to apply our models there is no need to manually distinguish the accelerometer data between walking and running, as in previously developed equations [@Neugebauer2014], which facilitates its use with daily living accelerometer data; and v) along with the creation of these models, a software was developed in the form of an R package to perform all the computations needed to get from the raw accelerometer data to the mechanical loading prediction.

@@

Despite the validity of the prediction models developed here, some limitations should be mentioned. First, although the activities included in our protocol (walking and running) are among the most frequently performed motor activities during daily living, our prediction models are not applicable to other activities known to influence bone adaptation to mechanical stimuli, such as jumping and direction changing. Second, in the absence of another sample for model validation, the leave-one-out cross-validation method was used, which is the method of choice in these situations [@Staudenmayer2012]. Nevertheless, future studies should test the out-of-sample validity of these models. Also, the protocol was performed on an instrumented treadmill. The use of a treadmill, although enhancing the study internal validity by the continuously measuring the GRF [@Ngoh2018] also reduces the external validity due to possible biomechanical changes in the walking and running patterns induced by the treadmill [@Kluitenberg2012]. Lastly, bone adaptation to mechanical stimulation depends not only on external loads, such as GRF, but also on internal loads generated by muscle forces [@Robling2009]. As accelerometers cannot reliably assess muscle force generation, they may have a limited capacity in estimating the bone response to loading.

>>>>

```{r accuracy_no_cv}
accuracy_no_cv <- map_dfr(res_GRF_models, accuracy) %>%
  select(MAPE) %>%
  mutate(MAPE = as.numeric(MAPE))
accuracy_no_cv <- myround(max(accuracy_no_cv) * 100, 1)
```

@@

Data from FP were collected at 1000 Hz and from accelerometers at 100 Hz, which is the GT9X Link maximum sampling frequency, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and acceleration (ACC; expressed in gravitational ACC units; 1*g* = 9.807 m·s^-2^) from the x, y and z vectors were exported to plain text files.

@@

FP and accelerometer data were processed using MATLAB (Version 2019a, Mathworks, Natick, USA), as described elsewhere [@Veras2020]. Briefly, both GRF and acceleration signals were filtered using a Butterworth fourth-order low-pass filter, with 20 Hz cut-off frequency, and their resultant vector calculated ($r_i = \sqrt{x_i^2 + y_i^2 + z_i^2}$). After that, peak ACC (pACC) values were defined as minimum height of twice the positive ACC average, during any given speed, and separated by 0.4 seconds minimum. GRF and ACC signals were adjusted through the time set by the systems clock, with manual correction based on visual inspection for possible errors, and then synchronized by the maximum cross-correlation coefficient. Peak GRF (pGRF) was determined as the highest value between 0.4 seconds before and after each pACC. The rates of change were computed through a centered derivative from the beginning of the foot contact to the peak (Equation 1). The peak loading rate (pLR) and peak acceleration rate (pAR) were defined as the maximum value in the array (Equation 2).

@@

Finally, our pGRF prediction equation was compared with a previously published reference equation [@Neugebauer2014], that was developed using similar methods. This comparison was performed in three ways using: i) the whole sample; ii) a subsample of walking trials; iii) a subsample of running trials. To assess the prediction accuracy, MAE, MAPE and RMSE were calculated using pGRF values predicted from both equations. No pLR prediction equation developed for both walking and running was found to be compared with ours.

@@

Furthermore, we compared the performance of our equations with a reference equation previously published [@Neugebauer2014]. As the only comparable equation found was the one to predict pGRF vertical vector using hip-worn accelerometer data developed by Neugebauer et al [@Neugebauer2014], we applied it to our data and computed the MAE, MAPE and RMSE. Then, these indices were compared to the indices for our equation (Table 1). Apart from the overall equations performance, a sub-analysis was done separating walking and running trials. The results of these comparisons can be observed in the Table 2. Accuracy indices of our equations were all better compared to the reference equation. Furthermore, our equation performance was constant between walking and running, indicating that it can be used for both activities without loss of accuracy, which cannot be observed in the reference equation.

@@

Comparing our results with previous studies that used a similar approach to predict pGRF, it can be seen that prediction errors from our models are comparable with some studies whose best results presented a MAPE of approximately 6.0\% but which included only walking in their protocol [@Fortune2014; @Neugebauer2018; @Veras2020]. Additionally, previous studies that developed pGRF prediction models during both walking and running showed a MAPE of 8.3\% [@Neugebauer2014], which is higher than our results, and 5.2\% [@Neugebauer2012] with no model cross-validation, which is similar to our results when no cross-validation method is applied (highest value of `r accuracy_no_cv`\% in our models). Nevertheless, models from both studies present a caveat, that is the inclusion of a term in their equations that distinguishes whether the activity is walking or running, which is not easy to be identified from free-living accelerometer data, hindering the real-world application of these previous equations. Also, one of these studies included in their sample only children and adolescents [@Neugebauer2012], that present a higher variability in their gait pattern compared to the adults [@Gouelle2016], making this equation unsuitable to use in an adult population. Furthermore, it was observed that the reference equation [@Neugebauer2014] used to compare with ours has a much lower accuracy during walking compared with running (16.6\% vs. 9.7\%). As this reference equation was not validated in a sample including people with obesitu, it performs poorly when applied to data from people with high body mass, as is the case in our sample that performed the walking trials, compared with running trials mostly performed by non-obese people. It is also important to mention that the prediction error found in our pGRF models is lower than the error typically found in models developed to estimate energy expenditure [@Diniz-Sousa2020], which is one of the main uses of accelerometer data [@Migueles2017].

@@

To the best of our knowledge, only two studies have already attempted to develop pLR prediction models based on accelerometry data, and both studies only included walking activities. Accordingly to what we found, these studies also showed that the prediction error for these models is considerably larger than for the pGRF prediction. One of these researches investigated the use of four different accelerometer placements (ankle, tibia, thigh and waist) to predict pLR and achieved a MAPE ranging from 23.4\% to 46.1\% [@Fortune2014], while the other obtained values from 19.5\% to 22.4\% using accelerometers at the lower back and hip [@Veras2020]. Results from the models developed in these previous investigations have both lower and higher prediction errors compared to our results, which were approximately of 35\%. Although pLR is an important variable, known for a long time to determine to a great extent the bone formation response to mechanical loading [@Bassey1998; @Turner1995], based on our findings and in previous studies [@Fortune2014; @Veras2020], its prediction through accelerometer data does not seem to yield trustworthy results.
\medskip

Results from our study offer some advantages compared to previous investigations: i) it includes walking and running, which are among the most usual types of physical activity performed daily; ii) all models developed, especially for pGRF prediction, presented a similar or better accuracy compared to the results from previous publications; iii) the accuracy of our models is fairly constant throughout walking and running, notably for the pGRF prediction models, with a MAPE increase from 6\% to 8\%, showing, therefore, that they can be used adequately for both activities; iv) in order to apply our models there is no need to manually distinguish the accelerometer data between walking and running, as in previously developed equations [@Neugebauer2014], which facilitates its use with daily living accelerometer data; and v) along with the creation of these models, a software was developed in the form of an R package to perform all the computations needed to get from the raw accelerometer data to the mechanical loading prediction.

@@

Despite the validity of the prediction models developed here, some limitations should be mentioned. First, although the activities included in our protocol (walking and running) are among the most frequently performed motor activities during daily living, our prediction models are not applicable to other activities known to influence bone adaptation to mechanical stimuli, such as jumping and direction changing. Second, in the absence of another sample for model validation, the leave-one-out cross-validation method was used, which is the method of choice in these situations [@Staudenmayer2012]. Nevertheless, future studies should test the out-of-sample validity of these models. Also, the protocol was performed on an instrumented treadmill. The use of a treadmill, although enhancing the study internal validity by the continuously measuring the GRF [@Ngoh2018] also reduces the external validity due to possible biomechanical changes in the walking and running patterns induced by the treadmill [@Kluitenberg2012]. Lastly, bone adaptation to mechanical stimulation depends not only on external loads, such as GRF, but also on internal loads generated by muscle forces [@Robling2009]. As accelerometers cannot reliably assess muscle force generation, they may have a limited capacity in estimating the bone response to loading in situations where there are no major gravitational loadings involved.
